An efficient contact value approach to approximating the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-07-17 papers/pair-correlation/paper.tex
edits from Paho Tested on bennet. Took 0:19.71.

In a previous paper~\cite{schulte2012using}, we introduce a
functional that gives a good approximation for the pair distribution
function averaged over positions $\rr_2$ that are in contact with
$\rr_1$, defined as:
\begin{align}
  g_\sigma(\rr_1) &\equiv \frac{ \int g^{(2)}(\rr_1,\rr_2) \delta(\sigma -|\rr_1-\rr_2|)n(\rr_2)
    d\rr_2 }{ \tilde{n}(\rr_1)  }
\end{align}
where the weighted density $\tilde{n}(\rr_1)$ is defined by:
\begin{align}
  \tilde{n}(\rr) &\equiv \int n(\rr') \delta(\sigma -|\rr - \rr'|)d\rr'.
\end{align}
In \cite{schulte2012using} we use the contact-value theorem to derive the exact formula:
\begin{align}
  g_\sigma(\rr)% &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac12 \frac{1}{k_BT n(\rr) \tilde{n}(\rr)} \frac{\delta
    F_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:gsigma-exact}
\end{align}
where $\sigma(\rr)$ is the diameter of hard spheres located at
position $\rr$, and $F_{HS}$ is the Helmholtz free energy of the
hard-sphere fluid.  The functional derivative of the free energy with
respect to hard-sphere diameter in Eq.~\ref{eq:gsigma-exact} requires
that we be able to evaluate the change in free energy resulting from a
change in the diameter of specifically the hard spheres located at
position $\rr$.  This somewhat unusual construction is mathematically
straightforward within Fundamental Measure Theory
(FMT)~\cite{rosenfeld1989free}.  We employ the White Bear variation of
the FMT free energy functional~\cite{roth2002whitebear}, which
provides an excellent approximation for this averaged value of the
pair distribution function at contact for a variety of interfaces, and
over a wide range of densities.

@@

\section{Making the CVA efficient}
The existing approaches to approximating the pair distribution
function outlined in Section~\ref{sec:gn} have not been widely used in
the construction of density functionals based on thermodynamic
perturbation theory, largely due to their computational complexity.
While our CVA provides only an incremental improvement in accuracy,
its construction enables significant gains in computational
efficiency, allowing for practical application in density functionals.
We achieve this gain by developing a \emph{separable} fit to the radial
distribution function of the hard-sphere fluid (see
Section~\ref{sec:separable-fit} for details).  This separable fit is
of the form
\begin{align}
  g(|r_{12}|; g_\sigma) &= \sum_{i} a_i(r_{12}) b_i(g_\sigma)
\end{align}
where the notable aspect is that the radial distribution function is
written as a sum of terms that are each a simple product of a function
of radius with a function of $g_\sigma$.  This enables us to write
integrals---such as Eq.~\ref{eq:mean-field}---that are linear in the
pair distribution function as a summation of fixed-kernel
convolutions, which may be efficiently computed using Fast Fourier
Transforms (FFTs) in $\mathcal{O}(N\log N)$ time.

@@

We begin by examining the lowest-order correction in TPT, given by
Eq.~\ref{eq:mean-field}.  The two terms that are averaged in
Eq.~\ref{eq:g2-our-mean} give equal contributions to the integral
\begin{align}
  F_1^{\textit{CVA}} &= \tfrac12 \!\! \iint \!\!
  g(r_{12};g_\sigma(\rr_2))n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|)
  d\rr_1d\rr_2.
\end{align}
When we introduce the separable form for $g(r_{12};g_\sigma)$ we can
further simplify this integral as
\begin{align}
  F_1^{\textit{CVA}} \! = \!
  \sum_i \!\tfrac12 \!\! \int \!\! n(\rr_1) \!\!
                            \int \!\!\!
                            %\big(
                            a_i(r_{12})\Phi(r_{12})
                            %\big)
                            %\big(
                            b_i(
g_\sigma(\rr_2))n(\rr_2)
                            %\big)
  d\rr_2d\rr_1
\end{align}
where the functional is written as a summation of integrals, each of
which is an integral of a simple convolution in three dimensions.  Thus,
each of these integrals may be computed in $\mathcal{O}(N\log N)$
time, which is the same scaling behavior that is common to the most
common and accurate classical density functional theories, most
notably Fundamental Measure Theory (FMT)~\cite{rosenfeld1989free,
  roth2002whitebear}.

@@

Having settled on the basic structure of our function, we further
refine it by performing a separable fit to the radial distribution
function from Monte Carlo simulation.  We chose to focus our fit on
the range of distances $r_{12} \le \maxrfit R$.  This range is
relevant to the widely used \cite{chapman1989saft,
  muller2001molecular, tan2008recent} Statistical
Associating Fluid Theory of Variable Range (SAFT-VR) free energy with
square-well dispersive attraction developed by Gil-Villegas \emph{et
  al.}~\cite{gil1997statistical}.
%
Although we consider this range of radii particularly interesting,
this is not a fundamental limit of the approach, as one could readily
extend the fit to larger radii by including additional fitting
parameters.
%
For comparison, in Fig.~\ref{fig:radial-distribution} we plot our fit,
Monte-Carlo data, and the radial distribution function of Gil-Villegas
\emph{et al.}, which we have extracted from their approximation for
the first term in the dispersion free energy given by
Eq.~\ref{eq:mean-field}.

For ease of implementation and future extension to larger radii, we
chose to express our fit with a fourth-order polynomial.  We constrain
our functional form such that $g(r; g_\sigma)$ reduces to $g_\sigma$
at contact and approaches $g(r)=1$ in the low-density limit.
Incorporating these constraints we have the functional form
\begin{align}
  g(r;g_\sigma) &=
  g_\sigma + \sum_{i=1}^{4} \sum_{j=1}^{4} \kappa_{ij} (g_\sigma - 1)^i
  \left(\tfrac{r}{\sigma}-1\right)^j,
  \label{eq:fit-form}
\end{align}
where the matrix $\kappa_{ij}$ is determined from a least-squares
fit to Monte Carlo data for the radial distribution function, over the
range $2R \le r \le \maxrfit R$, and for packing fractions $\eta \le
0.45$.  The resulting parameters are displayed in
Table~\ref{tab:kappa}.  The maximum error in $g(r)$ within this
range is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr R$.  Fig.~\ref{fig:radial-distribution} displays
our approximation at just under half of the densities that were
included in the fit.

>>>>

In a previous paper~\cite{schulte2012using}, we introduce a
functional that gives a good approximation for the pair distribution
function averaged over positions $\rr_2$ that are in contact with
$\rr_1$, defined as:
\begin{align}
  g_\sigma(\rr_1) &\equiv \frac{ \int g^{(2)}(\rr_1,\rr_2) \delta(\sigma -|\rr_1-\rr_2|)n(\rr_2)
    d\rr_2 }{ \tilde{n}(\rr_1)  }
\end{align}
where the weighted density $\tilde{n}(\rr_1)$ is defined by:
\begin{align}
  \tilde{n}(\rr) &\equiv \int n(\rr') \delta(\sigma -|\rr - \rr'|)d\rr'.
\end{align}
In \cite{schulte2012using} we use the contact-value theorem to derive the exact formula:
\begin{align}
  g_\sigma(\rr)% &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac12 \frac{1}{k_BT n(\rr) \tilde{n}(\rr)} \frac{\delta
    F_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:gsigma-exact}
\end{align}
where $\sigma(\rr)$ is the diameter of hard spheres located at
position $\rr$, and $F_{HS}$ is the Helmholtz free energy of the
hard-sphere fluid.  The functional derivative of the free energy with
respect to the hard-sphere diameter in Eq.~\ref{eq:gsigma-exact} requires
that we be able to evaluate the change in free energy resulting from a
change in the diameter of specifically the hard spheres located at
position $\rr$.  This somewhat unusual construction is mathematically
straightforward within Fundamental Measure Theory
(FMT)~\cite{rosenfeld1989free}.  We employ the White Bear variation of
the FMT free energy functional~\cite{roth2002whitebear}, which
provides an excellent approximation for this averaged value of the
pair distribution function at contact for a variety of interfaces, and
over a wide range of densities.

@@

\section{Making the CVA efficient}
The existing approaches to approximating the pair distribution
function outlined in Section~\ref{sec:gn} have not been widely used in
the construction of density functionals based on thermodynamic
perturbation theory, largely due to their computational complexity.
While our CVA provides only an incremental improvement in accuracy,
its construction enables significant gains in computational
efficiency, allowing for practical application in density functionals.
We achieve this gain by developing a \emph{separable} fit to the radial
distribution function of the hard-sphere fluid (see
Section~\ref{sec:separable-fit} for details).  This separable fit is
of the form
\begin{align}
  g(r_{12}; g_\sigma) &= \sum_{i} a_i(r_{12}) b_i(g_\sigma)
\end{align}
where the notable aspect is that the radial distribution function is
written as a sum of terms that are each a simple product of a function
of radius with a function of $g_\sigma$.  This enables us to write
integrals---such as Eq.~\ref{eq:mean-field}---that are linear in the
pair distribution function as a summation of fixed-kernel
convolutions, which may be efficiently computed using Fast Fourier
Transforms (FFTs) in $\mathcal{O}(N\log N)$ time.

@@

We begin by examining the lowest-order correction in TPT, given by
Eq.~\ref{eq:mean-field}.  The two terms that are averaged in
Eq.~\ref{eq:g2-our-mean} give equal contributions to the integral
\begin{align}
  F_1^{\textit{CVA}} &= \tfrac12 \!\! \iint \!\!
  g(r_{12};g_\sigma(\rr_2))n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|)
  d\rr_1d\rr_2.
\end{align}
When we introduce the separable form for $g(r_{12};g_\sigma)$ we can
further simplify this integral as
\begin{align}
  F_1^{\textit{CVA}} \! = \!
  \sum_i \!\tfrac12 \!\! \int \!\! n(\rr_1) \!\!
                            \int \!\!\!
                            %\big(
                            a_i(r_{12})\Phi(r_{12})
                            %\big)
                            %\big(
                            b_i(
g_\sigma(\rr_2))n(\rr_2)
                            %\big)
  d\rr_2d\rr_1
\end{align}
where the functional is written as a summation of integrals of simple
convolutions in three dimensions.  Thus,
each of these integrals may be computed in $\mathcal{O}(N\log N)$
time, which is the same scaling behavior that is common to the most
common and accurate classical density functional theories, most
notably Fundamental Measure Theory (FMT)~\cite{rosenfeld1989free,
  roth2002whitebear}.

@@

Having settled on the basic structure of our function, we further
refine it by performing a separable fit to the radial distribution
function from Monte Carlo simulation.  We focus our fit on
the range of distances $r_{12} \le \maxrfit R$.  This range is
relevant to the widely used \cite{chapman1989saft,
  muller2001molecular, tan2008recent} Statistical
Associating Fluid Theory of Variable Range (SAFT-VR) free energy with
square-well dispersive attraction developed by Gil-Villegas \emph{et
  al.}~\cite{gil1997statistical}.
%
Although we consider this range of radii particularly interesting,
this is not a fundamental limit of the approach, as one could readily
extend the fit to larger radii by including additional fitting
parameters.
%
For comparison, in Fig.~\ref{fig:radial-distribution} we plot our fit,
Monte-Carlo data, and the radial distribution function of Gil-Villegas
\emph{et al.}, which we have extracted from their approximation for
the first term in the dispersion free energy given by
Eq.~\ref{eq:mean-field}.

For ease of implementation and future extension to larger radii, we
fit the radial distribution function using a fourth-order polynomial.  We constrain
our functional form such that $g(r; g_\sigma)$ reduces to $g_\sigma$
at contact and approaches $g(r)=1$ in the low-density limit.
Incorporating these constraints we have the functional form
\begin{align}
  g(r;g_\sigma) &=
  g_\sigma + \sum_{i=1}^{4} \sum_{j=1}^{4} \kappa_{ij} (g_\sigma - 1)^i
  \left(\tfrac{r}{\sigma}-1\right)^j,
  \label{eq:fit-form}
\end{align}
where the matrix $\kappa_{ij}$ is determined from a least-squares
fit to Monte Carlo data for the radial distribution function, over the
range $2R \le r \le \maxrfit R$, and for packing fractions $\eta \le
0.45$.  The resulting parameters are displayed in
Table~\ref{tab:kappa}.  The maximum error in $g(r)$ within this
range is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr R$.  Fig.~\ref{fig:radial-distribution} displays
our approximation at just under half of the densities that were
included in the fit.
